Estimating the dynamics of \\SPARQL query results

### 2019-04-11 kgswc19.tex

@@

\begin{document}
%
\title{Estimating the dynamics of SPARQL query results
%\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}
}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{Alberto Moya\orcidID{0000-0002-7003-5087} \and 
%Aidan Hogan\orcidID{0000-0001-9482-1982}}
%
%\authorrunning{A. Moya and A. Hogan}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{IMFD; DCC, University of Chile\\
%\email{\{amoya,ahogan\}@dcc.uhile.cl}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We address the problem of estimating when the results of an input SPARQL query results over dynamic RDF datasets will change. We evaluate a framework that extracts features from the query and/or from past versions of the target dataset. The resulting features are then used to predict (1) whether or not the results for a query will change in the near future using classifiers, and (2) when they might change (if at all) using linear regression. To evaluate, we create a gold standard based on 23 versions of Wikidata and a curated collection of 221 SPARQL queries. Our results show that the quality of predictions using query features and lightweight statistics of the dynamics of individual predicates is less competitive with that obtained using (more costly to derive) knowledge of the complete historical changes in the query results.

@@

Some works have studied the data dynamics based on the historical evolution of entities~\cite{KaferAUOH13,DividinoGSG14,NishiokaS15}, but following such an approach for SPARQL queries is expensive because (1) a SPARQL query may involve potentially many entities; and (2) it is necessary to have the previous complete versions of data to analyse the entities relevant to a query. Many works have explored the dynamics of Linked Data with the intention of finding patterns that allow for characterizing, recognizing, and predicting changes based on analysis of the domains, predicates, and schema~\cite{UmbrichHHPD10,KaferAUOH13,NishiokaS16,NishiokaS17,GonzalezH18}. Among these works, we can find hybrid approaches that have been developed to return fresher query results with faster response times by decomposing a query into dynamic sub-queries executed remotely, and static sub-queries executed over local caches~\cite{UmbrichKHP12}, but such an approach only focuses on estimating the dynamics of individual triple patterns, and does not consider, for example, query operators. More generally, coping with changes in remote data still presents a major challenge for applications leveraging dynamic Linked Data.

@@

\paragraph{Paper structure:} In Section~\ref{sec:related}, we review related works. Section~\ref{sec:preliminar} provides a formal statement of the problem. Section~\ref{sec:approach} describes the evaluated architecture and features. In Section~\ref{sec:data}, we describe the creation of our gold standard based on Wikidata. Section~\ref{sec:eval} presents the design and results of the experiments. We conclude and review future directions in Section~\ref{sec:conclusion}.

@@

\paragraph{Query Features:} Our initial set of features are based on analysis of the input SPARQL query. Figure~\ref{fig:query} shows only a sample of the evaluated features, as well as -- for the purposes of illustration -- their values for an accompanying example query.

@@

Although we have seen that the number of unique predicates did not change much over all versions, we are interested to see which predicates are indicative of dynamic statements. Table~\ref{tab:predicates} shows the ten most dynamic predicates according to the number of added and deleted statements involving that predicate (Total) and the ratio of added and deleted statements divided by the total number of statements for that predicate across all snapshots (Dyn); for the second, we only include predicates that appear in all snapshots and appear in ≥ 1, 000 statements overall. Note that the largest number of changes involves the predicate scheme:description, however, its dynamic value is low due to the number of statements with it in the versions.

@@

96 queries had to be removed from the original set of 389 because they asked for information external to Wikidata (using \texttt{SERVICE}) or asked for qualifiers that are not present in the truthy version. We also eliminated 12 queries that returned bindings to blank nodes to facilitate comparison between the results. Furthermore, some queries featured non-deterministic elements -- such as \texttt{LIMIT}/\texttt{OFFSET} without ordering, \texttt{SAMPLE} or temporal functions -- such as \texttt{TODAY()} -- that may lead to changes in results not related to changes in the data; we remove 23 queries with \texttt{SAMPLE} and temporal functions, and add an \texttt{ORDER BY} clause to all queries to ensure determinism with \texttt{LIMIT}/\texttt{OFFSET} and more generally to facilitate quick comparison of results. We also eliminated 37 queries with 0 results in the executions over all the snapshots. As a result of this process of filtering, we end up with a total of 221 deterministic queries.

Furthermore, we needed to transform certain queries for the purposes of our experiments. 163 queries used features particular to the Wikidata Query Service, such as using a custom label \texttt{SERVICE} call to specify language preferences; we converted these queries to standard SPARQL using \texttt{COALESCE} to capture preferences.  Figure~\ref{fig:transform} shows an example of a Wikidata query with the \texttt{LABEL} service rewritten and a deterministic ordering added.

\begin{figure}[!t]
\centering
{
	\begin{minipage}{0.8\textwidth}
	\begin{lstlisting}[style=sparqld]
#Cats
SELECT ?item ?itemLabel
WHERE {
 ?item wdt:P31 wd:Q146 .
 SERVICE wikibase:label { bd:serviceParam wikibase:language "en,es" }
}
	\end{lstlisting}
	\end{minipage}	
}

{
\begin{minipage}{0.8\textwidth}
	\begin{lstlisting}[style=sparqld]
#Cats
SELECT ?item ?itemLabel
WHERE {
  ?item wdt:P31 wd:Q146 .
  OPTIONAL { ?item rdfs:label ?en . FILTER(lang(?en)="en") }
  OPTIONAL { ?item rdfs:label ?es . FILTER(lang(?es)="es") }
  BIND(str(coalesce(?en, ?es, strafter(str(?item),
    "http://www.wikidata.org/entity/"))) AS ?itemLabel )
}
ORDER BY ?item ?itemLabel
	\end{lstlisting}
\end{minipage}	
}
\caption{Query transformation before (above) and after (below) \label{fig:transform}}
\end{figure}

@@

\paragraph{Setting:} To build our final datasets, given that our data features are temporally based, we must thus define a past interval to consider. With 23 versions, we must hold out at least one version to label, leaving a maximum interval of 22 past versions to use to compute our data features. However, the more past versions we consider for a prediction, the fewer examples we can generate; for example, if we consider 22 versions to compute, we will only have one ground truth label for each query. In the end, we opted to consider intervals of 3, 5, 9 and 17 previous versions, for example, ($G_{i-3}, G_{i-2}, G_{i-1}$) to predict changes in queries results for the subsequent version ($G_i$); this allows us to predict from $G_4$ to $G_{23}$ inclusive, providing 20 examples per query and 4420 examples in the largest dataset and 1326 examples in the smallest dataset. We remark that no feature explicitly indicates which specific version numbers are given and which are to be predicted.

@@

In addition, we train a Machine Learning model to determine the TTL of a query's results. To train the model we consider the changes observed in the monitored interval. For each update of the results of a query, we have created a training example with the previously presented features. The target TTL value that we learn is the time interval during which the results of the query remain fresh from the previous update time. As an example, suppose that the query q change three times ($G_{3}, G_{7}, G_{9}$) over all versions, we consider two examples from $G_{3}$ and $G_{7}$ with target TTL value 4 and 2 respectively.

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-score for the predictions made considering the union of different combinations of the sets of features (Q, P, R) and varying the window sizes (3, 5, 9, 17); for reference, we include a baseline that randomly guesses yes/no. 
We can note the best results were obtained using the features from \textsc{r} reaching values ​​of 0.83 of $F_1$. The results show that predicting changes in query results is indeed a challenging problem. Looking in further detail, as can be expected, prediction performs best when the \textsc{r} feature - with knowledge of changes in the historical value of results of the input query - is available. And, the features from sets \textsc{q} and  \textsc{p} - which do not assume the availability of such information - obtain less $F_1$-score: more specifically, the best result was using \textsc{pr} (0.83) with Linear SVM and window size 17, while the best results without \textsc{r} was using \textsc{Q} (0.60) with Linear SVM and window size 5.

Furthermore, we show in Figure~\ref{fig:result} that effectively the size of the window influences the quality of the results when considering \textsc{r} (blue), being better as the window increases. However, these results are not affected when we consider only predicates in \textsc{q} (brown), \textsc{p} (green)  or \textsc{qp} (black). Also, we show, that all configurations outperform the random baseline (red).

@@

		& \textbf{Better}          & \textbf{0.596} & \textbf{0.519} & \textbf{0.781} & \textbf{0.593} & \textbf{0.775} & \textbf{0.781} & \textbf{0.774} \\    \bottomrule

@@

		& \textbf{Better}          & \textbf{0.6}   & \textbf{0.535} & \textbf{0.795} & \textbf{0.59}  & \textbf{0.787} & \textbf{0.797} & \textbf{0.787} \\    \bottomrule

@@

		& \textbf{Better}          & \textbf{0.578} & \textbf{0.528} & \textbf{0.797} & \textbf{0.571} & \textbf{0.795} & \textbf{0.8}   & \textbf{0.796} \\    \bottomrule

@@

\begin{figure}[t]
	%\begin{minipage}[b]{0.30\linewidth}
	\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={F1},
	width=5cm, 
	height=5cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

	\end{axis}
	\end{tikzpicture}
	%		\caption{Evolution of the number of triples on Wikidata.}
	%		\label{graph:datasize}
	%	\end{minipage}
	\hspace{0.1cm}
	%	\begin{minipage}[b]{0.30\linewidth}
	%		\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={Precision},
	width=5cm, 
	height=5cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

	\end{axis}
	\end{tikzpicture}
	%		\caption{Evolution of the number of triples on Wikidata.}
	%		\label{graph:datasize}
	%	\end{minipage}
	\hspace{0.1cm}
	%	\begin{minipage}[b]{0.30\linewidth}
	%		\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={Recall},
	width=5cm, 
	height=5cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

Our goal then was to explore this trade-off experimentally. For this, we define some desirable characteristics for the queries and data used in the experimental setting, such as availability, size, dynamics, diversity, etc. Settling on Wikidata, we collect 23 versions and curate 221 user-generated queries from which we compute results, statistics and ultimately features that form the gold standard for the tasks. Analysing these data and queries, we conclude that they meet the desired characteristics, offering a large-scale, diverse, real-world setting; more importantly, we find that our queries offer an ideal distribution of queries whose results always change, queries whose results never change, and (more importantly) queries whose results change intermittently and for which prediction is thus a non-trivial task. We make this gold standard available at \ah{...} -- including queries, results, and features -- so that the results presented here can be reproduced and so that it can be reused in further research.

>>>>

\makeatletter \renewcommand\subsubsection{\@startsection{subsubsection}{4}{\z@}%
	{1ex \@plus0ex \@minus0ex}% 
	{-1ex}% 
	{\normalfont\normalsize\bfseries}} 
\makeatother

@@

\begin{document}
%
\title{Estimating the dynamics of \\SPARQL query results
%\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}
}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{Alberto Moya\orcidID{0000-0002-7003-5087} \and 
%Aidan Hogan\orcidID{0000-0001-9482-1982}}
%
%\authorrunning{A. Moya and A. Hogan}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{IMFD; DCC, University of Chile\\
%\email{\{amoya,ahogan\}@dcc.uhile.cl}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We address the problem of estimating when the results of an input SPARQL query results over dynamic RDF datasets will change. We evaluate a framework that extracts features from the query and/or from past versions of the target dataset. The resulting features are then used to predict (1) whether or not the results for a query will change in the near future using classifiers, and (2) when they might change (if at all) using linear regression. To evaluate, we create a gold standard based on 23 versions of Wikidata and a curated collection of 221 SPARQL queries. Our results show that the quality of predictions using query features and lightweight statistics of the dynamics of individual predicates is less competitive with that obtained using (more costly to derive) knowledge of the complete historical changes in the query results.

@@

Some works study data dynamics based on the historical evolution of entities~\cite{KaferAUOH13,DividinoGSG14,NishiokaS15}, but following such an approach for SPARQL queries is expensive because (1) a SPARQL query may involve potentially many entities; and (2) it is necessary to have the previous complete versions of data to analyse the entities relevant to a query. Many works have explored the dynamics of Linked Data with the intention of finding patterns that allow for characterizing, recognizing, and predicting changes based on analysis of the domains, predicates, and schema~\cite{UmbrichHHPD10,KaferAUOH13,NishiokaS16,NishiokaS17,GonzalezH18}. Among these works are hybrid approaches developed to return fresher query results with faster response times by decomposing a query into dynamic sub-queries executed remotely, and static sub-queries executed over local caches~\cite{UmbrichKHP12}, but such an approach only focuses on estimating the dynamics of individual triple patterns, and does not consider, for example, query operators. More generally, coping with changes in remote data still presents a major challenge for applications leveraging dynamic Linked Data.

@@

%\paragraph{Paper structure:} In Section~\ref{sec:related}, we review related works. Section~\ref{sec:preliminar} provides a formal statement of the problem. Section~\ref{sec:approach} describes the evaluated architecture and features. In Section~\ref{sec:data}, we describe the creation of our gold standard based on Wikidata. Section~\ref{sec:eval} presents the design and results of the experiments. We conclude and review future directions in Section~\ref{sec:conclusion}.

@@

\paragraph{Query Features:} Our initial set of features are based on analysis of the input SPARQL query. Figure~\ref{fig:query} shows only a sample of evaluated features, as well as -- for the purposes of illustration -- their values for an example query.

@@

Although we have seen that the number of unique predicates did not change much over all versions, we are interested to see which predicates are indicative of dynamic statements. Table~\ref{tab:predicates} shows the ten most dynamic predicates according to the number of added and deleted statements involving that predicate (Total) and the ratio of added and deleted statements divided by the total number of statements for that predicate across all snapshots (Dyn); for the second, we only include predicates that appear in all snapshots and appear in $\geq$ 1,000 statements overall. Note that the largest number of changes involves the predicate scheme:description, however, its dynamic value is low due to the number of statements with it in the versions.

@@

A total of 96 queries had to be removed from the original set of 389 because they asked for information external to Wikidata (using \texttt{SERVICE}) or asked for qualifiers that are not present in the truthy version. We also eliminated 12 queries that returned bindings to blank nodes to facilitate comparison between the results. Furthermore, some queries featured non-deterministic elements -- such as \texttt{LIMIT}/\texttt{OFFSET} without ordering, \texttt{SAMPLE} or temporal functions -- such as \texttt{TODAY()} -- that may lead to changes in results not related to changes in the data; we remove 23 queries with \texttt{SAMPLE} and temporal functions, and add an \texttt{ORDER BY} clause to all queries to ensure determinism with \texttt{LIMIT}/\texttt{OFFSET} and more generally to facilitate quick comparison of results. We also eliminated 37 queries with 0 results forr all the snapshots. As a result of this process of filtering, we end up with a total of 221 deterministic queries.

%Furthermore, we needed to transform certain queries for the purposes of our experiments. 163 queries used features particular to the Wikidata Query Service, such as using a custom label \texttt{SERVICE} call to specify language preferences; we converted these queries to standard SPARQL using \texttt{COALESCE} to capture preferences.  Figure~\ref{fig:transform} shows an example of a Wikidata query with the \texttt{LABEL} service rewritten and a deterministic ordering added. 
%
%\begin{figure}[!t]
%\centering
%{
%	\begin{minipage}{0.8\textwidth}
%	\begin{lstlisting}[style=sparqld]
%#Cats
%SELECT ?item ?itemLabel
%WHERE {
% ?item wdt:P31 wd:Q146 .
% SERVICE wikibase:label { bd:serviceParam wikibase:language "en,es" }
%}
%	\end{lstlisting}
%	\end{minipage}	
%}
%
%{
%\begin{minipage}{0.8\textwidth}
%	\begin{lstlisting}[style=sparqld]
%#Cats
%SELECT ?item ?itemLabel
%WHERE {
%  ?item wdt:P31 wd:Q146 .
%  OPTIONAL { ?item rdfs:label ?en . FILTER(lang(?en)="en") }
%  OPTIONAL { ?item rdfs:label ?es . FILTER(lang(?es)="es") }
%  BIND(str(coalesce(?en, ?es, strafter(str(?item),
%    "http://www.wikidata.org/entity/"))) AS ?itemLabel )
%}
%ORDER BY ?item ?itemLabel
%	\end{lstlisting}
%\end{minipage}	
%}
%\caption{Query transformation before (above) and after (below) \label{fig:transform}}
%\end{figure}

@@

\paragraph{Setting:} To build our final datasets, we must define a past interval to consider. With 23 versions, we must hold out at least one version to label, leaving a maximum interval of 22 past versions to use to compute our data features. However, the more past versions we consider for a prediction, the fewer examples we can generate; for example, if we consider 22 versions, we will only have one ground truth label for each query. In the end, we opted to consider intervals of 3, 5, 9 and 17 previous versions, for example, ($G_{i-3}, G_{i-2}, G_{i-1}$) to predict changes in queries results for the subsequent version ($G_i$); this allows us to predict from $G_4$ to $G_{23}$ inclusive, providing 20 examples per query, 4,420 examples in the largest dataset and 1,326 examples in the smallest dataset. No feature explicitly indicates which specific version numbers are given and which are to be predicted.

@@

In addition, we train a Machine Learning model to determine the TTL of a query's results. To train the model we consider the changes observed in the monitored interval. For each update of the results of a query, we have created a training example with the previously presented features. The target TTL value that we learn is the time interval during which the results of the query remain fresh from the previous update time. As an example, suppose that the query $Q$ change three times ($G_{3}, G_{7}, G_{9}$) over all versions; we consider two examples from $G_{3}$ and $G_{7}$ with target TTL value 4 and 2 respectively.

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-score for the predictions made considering the union of different combinations of the sets of features (Q, P, R) and varying the window sizes (3, 5, 9, 17); for reference, we include a baseline that randomly guesses yes/no. 
We can note the best results were obtained using the features from \textsc{r} reaching values of 0.83 $F_1$. The results show that predicting changes in query results is indeed a challenging problem. Looking in further detail, as can be expected, prediction performs best when the \textsc{r} feature -- with knowledge of changes in the historical value of results of the input query -- is available. The features from sets \textsc{q} and  \textsc{p} -- which do not assume the availability of such information -- obtain a lower $F_1$-score: more specifically, the best result was using \textsc{pr} (0.83) with Linear SVM and window size 17, while the best results without \textsc{r} was using \textsc{Q} (0.60) with Linear SVM and window size 5.

Furthermore, we show in Figure~\ref{fig:result} that the size of the window influences the quality of the results when considering \textsc{r} (blue), being better as the window increases. However, these results are not affected when we consider only predicates in \textsc{q} (brown), \textsc{p} (green)  or \textsc{qp} (black). Also, we show, that all configurations outperform the random baseline (red).

@@

		& \textbf{Better}          & \textbf{0.596} & \textbf{0.519} & \textbf{0.781} & \textbf{0.593} & \textbf{0.775} & \textbf{0.781} & \textbf{0.774} \\    \midrule

@@

		& \textbf{Better}          & \textbf{0.6}   & \textbf{0.535} & \textbf{0.795} & \textbf{0.59}  & \textbf{0.787} & \textbf{0.797} & \textbf{0.787} \\    \midrule

@@

		& \textbf{Better}          & \textbf{0.578} & \textbf{0.528} & \textbf{0.797} & \textbf{0.571} & \textbf{0.795} & \textbf{0.8}   & \textbf{0.796} \\    \midrule

@@

\begin{figure}[t]
	%\begin{minipage}[b]{0.30\linewidth}
	\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={F1},
	width=5cm, 
	height=4cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

	\end{axis}
	\end{tikzpicture}
	%		\caption{Evolution of the number of triples on Wikidata.}
	%		\label{graph:datasize}
	%	\end{minipage}
	\hspace{0.1cm}
	%	\begin{minipage}[b]{0.30\linewidth}
	%		\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={Precision},
	width=5cm, 
	height=4cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

	\end{axis}
	\end{tikzpicture}
	%		\caption{Evolution of the number of triples on Wikidata.}
	%		\label{graph:datasize}
	%	\end{minipage}
	\hspace{0.1cm}
	%	\begin{minipage}[b]{0.30\linewidth}
	%		\centering
	\begin{tikzpicture}[thick,font=\scriptsize,scale=0.95]
	\begin{axis}[
	%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
	xlabel={Recall},
	width=5cm, 
	height=4cm,
	ymin=0,
	xtick={3,5,9,17},
	xticklabels={},
	extra x ticks={3,5,9,17},
	legend pos=north west,
	ymajorgrids=true,
	grid style=dashed,
	]

@@

We thus explore this trade-off experimentally. For this, we define some desirable characteristics for the queries and data used in the experimental setting, such as availability, size, dynamics, diversity, etc. Settling on Wikidata, we collect 23 versions and curate 221 user-generated queries from which we compute results, statistics and ultimately features that form the gold standard for the tasks. Analysing these data and queries, we conclude that they meet the desired characteristics, offering a large-scale, diverse, real-world setting; more importantly, we find that our queries offer an ideal distribution of queries whose results always change, queries whose results never change, and (more importantly) queries whose results change intermittently and for which prediction is thus a non-trivial task. We will make this gold standard available after double-blind review -- including queries, results, and features -- so that the results presented here can be reproduced and so that it can be reused in further research.
